---
title: "Climate reconstruction from tree rings by stepwise regression"
subtitle: "DISC 2026" 
execute:
  echo: false
format: 
  html:
    page-layout: full
    toc: true
    toc-location: right
    toc-title: Contents
theme: cerulean
bibliography: ReconStep.bib
---

The original Rmd file (ReconStep.Rmd) with full explanations, the UWFs it calls, sample input data, and running instructions are can be download from the [GitHub repository](https://github.com/ltrr-arizona-edu/TRISH-R)

```{r}
rm(list=ls())

pacman::p_load(here, rmarkdown, tidyverse, car, pracma, nortest, treeclim, resample, rjson, gplots, ggstackplot, ggthemes, patchwork)

# Edit following line to match where you have stored user-written functions (UWFs)
code_dir <- paste(here(), "/", sep = "")

source(paste(code_dir,'LagYear.R',sep=''))  # build lagged matrix of chrons
source(paste(code_dir,'PeriodCommon.R',sep=''))  # common period of overlap of chrons and predictand
source(paste(code_dir,'ForwStep1.R',sep='')) # forward stepwise regression
source(paste(code_dir,'mannken1.R',sep='')) # Mann-Kendall test for trend
source(paste(code_dir,'xyCI.R',sep='')) # utility function to compute x and y coordinates for patch for CI on recon plot
source(paste(code_dir,'ssValid.R',sep='')) # split-sample calib-valid
source(paste(code_dir,'NashSutt.R',sep='')) # independent verification check
source(paste(code_dir,'CrossValid1.R',sep='')) # cross-validation, called by ForwStep1

```

```{r}
clean_bib <- function(input_file, input_bib, output_bib){
  lines <- paste(readLines(input_file), collapse = "")
  entries <- unique(str_match_all(lines, "@([a-zA-Z0-9]+)[,\\. \\?\\!\\]\\;]")[[1]][, 2])

  bib <- paste(readLines(input_bib), collapse = "\n")
  bib <- unlist(strsplit(bib, "\n@"))

  output <- sapply(entries, grep, bib, value = T)
  output <- paste("@", output, sep = "")

  writeLines(unlist(output), output_bib)
}

# Call function to generate bib file for this project from my bib file. User
# can ignore this code. Be aware that clear_bib does not work properly with
# bibtex files, like mine, that have key fields with more than one name, such
# as "Meko&Stockton1983." So, I have to build ReconStep.bib MANUALLY pasting 
# from /media/dave/Red128/BibliographyMeko.bib to ReconStep.bib. To avoid catastropic 
# overwriting of ReconStep.bib, I name the NEVER-TO-BE-CREATED output file below
# 'ReconStepMeko.bib' instead of ReconStep.bib.
#
# clean_bib('ReconStep.Rmd','/home/dave/Data/BibTex/BibliographyMeko.bib','ReconStepMeko.bib')
```

```{r}
# Specify paths and filenames for the files with the input climate and tree-ring time series.
# A path, if included, must match the setup of your laptop. If no path is specified, the file
# is assumed to be in the R project directory.
# Edit following line (two lines) for path and filenames
fileClimate <- paste(here(), '/KGUPrecipWinter.txt', sep = "")
#'/home/dave/GitWork/TRISH-R/PsepAugTahoe.txt'
fileTree <- paste(here(), '/KGUChron.txt', sep = "")

# Specify desired calibration period of the reconstruction model. This period is also used for any 
# correlation screening of the pool of potential predictors, and for cross-validation of the
# reconstruction model. The specified first and last years apply to years of the y. Setting
# yrgoc and yrspc to "NA" means the calibration should be chosen automatically from the 
# maximum overlap of y (climate) and the tree-ring series. If you mistakenly specify an
# impossible calibration period, the settings will be ignored and the period will be 
# chosen automatically.
yrgoc<-NA  # start of calib period (if data and lags allow)
yrspc<- NA  # end of calib period (...)

# Set the period for "independent" verification. This must not overlap any of the years in
# yrgoc-yrspc. Independent verification checks how well the model predicts the predictand
# for a period NOT USED IN ANY WAY to screen or tune the predictand to predictors. If both
# yrgoi and yrspi are set to NA, no independent verification is attempted. This is an appropriate
# setting when there is no additional overlap of tree rings and the predictand beyond what
# is used for calibration/validation.
yrgoi <- NA # start year of independent verification
yrspi <- NA # end year of independent verification

# Specify the maximum number of negative and positive lags on the predictors allowed 
# in the predictor pool. The final model may not include such lags, depending on the 
# results of stepwise regression. The  maximum allowable lag is 2 years. In other 
# words, the only acceptable settings for nNeg and nPos are 0, 1, or 2.
nNeg=1     # maximum negative lag allowed on chronologies
nPos=1     # maximum positive lag allowed on chronologies

# Specify a decimal fraction that caps the size of the pool of potential predictors
# as some fraction of number of years in the calibration period. This is a safeguard
# against model overfitting. Regression models that select predictors automatically 
# can result in chance relationships if the pool of potential predictors is some large
# fraction of the number of observations (years) available for calibrating the model.
# Various rules exist on the allowable ratio of number of potential predictors to 
# number of observations for calibration. Ideally, there should be at least 5 to 10 
# times as many years in the calibration period as variables in the pool. Here by default
# the predictor pool is capped at no larger than 1/10 the number of observations in the
# calibration period. Beware that with lagged models the pool can become large even if
# using only a few chronologies. For example, 5 chronologies and a specified maximum lag
# of 2 years gives a pool of 25 potential predictors (five lags (-2 to +2) time 5
# chronologies. In that case, if the calibration period were 100 years, only the 10 
# potential predictors most highly correlated (Pearson) with the predictand would
# be allowed in the pool.
f<-1/10 # pool of potential predictors not allowed to be larger than this fraction of calibration period observations

# Optionally specify to transform y before regression using either a log10
# or square root transform. Keep in mind that log10 is invalid if y has 
# any zero or negative values, and that the square root transform is invalid
# if y has any negative values. 
#   1 = no transform
#   2 = square root transform
#   3 = log10 transform
ktransform<-1 # Transform predictand?  

# Specify the stopping rule for stepwise regression. Two options are available.
# Regardless of the option selected, regression proceeds stepwise forward, at
# each step entering the variable that gives the greatest reduction of residual
# variance from the previous step. The stopping rule specifies how the final 
# regression model is selected. One option is step at which the regression model
# has maximum adjusted R-squared. The other option is the step with maximum
# RE of cross-validation 
kStopWhen<-2  # 1=max adjusted R-squared, 2= maximum RE

# Specify whether the regression model for reconstruction uses original 
# chronologies or their principal components (PCs). A setting of "FALSE"
# instructs to use the screened lagged chronologies as the pool. A setting of
# "TRUE" instructs to use the PCs of those screened lagged chronologies.
PCA<-TRUE # regression on PCs of screened lagged chronologies

par(ask=TRUE) # pause before each plot; no need to edit this line

# END of user-defined variables/parameters
```

################################################################################ 

## PART 1. Read input, optionally transform *y*, and build a lagged matrix of chronologies

################################################################################ 

```{r}

# READ INPUT TIME SERIES
V <- read.table(fileClimate,sep="\t",header=FALSE) # input predictand, tab separated
#V <- read.table(fileClimate,sep="",header=FALSE) # input predictand, space separated
U <- read.table(fileTree,sep="\t",header=TRUE)  # input chronologies, tab-sep
#U <- read.table(fileTree,sep="",header=TRUE)  # input chronologies, space-sep

#--- TRIM PREDICTAND TO ELIMINATE NA'S
V<-as.matrix(V)
L<-complete.cases(V)
V<-V[L,]

#--- TRANSFORM PREDICTAND?
v<-V[,2]
yrv=V[,1]
if (ktransform==1){
  v<-v
  dunits <-'original'
} else if (ktransform==2) {
  v<-sqrt(v)
  dunits <- 'square root of original'
} else  {
  v=log10(v)
  dunits <- 'log10 of original'
}

 # want chronologies as matrix class
U<-as.matrix(U) 
yrU<-U[,1]
U<-U[,-1]
yrU<-as.matrix(yrU)
U<-as.matrix(U)

# Call a UWF to build a lagged matrix of the input chronologies. The columns of
# the lagged matrix, W, will be organized left to right in this order:
# 1) unlagged chronologies
# 2) negative lags on chronologies, beginning with lag t-1 from y
# 3) positive lags on chronologies, beginning with lag t+1 from y
mU<-nrow(U)
tgo<-yrU[1]
tsp<-yrU[mU]
ktrim=2
D<-LagYear(U,tgo,tsp,nNeg,nPos,ktrim) # list with lagged matrix, ids, etc
W<-D$X # lagged matrix of chrons
colnames(W)<-D$ids # add column names associating variables and lags with W
yrW<-D$tGo:D$tSp
# W, yrW, colnames(W) now desribed the full lagged chronology 

# Abort with error message if have called for reconstruction with PCA but 
# fewer than 3 variables in pool of potential predictors
if ((ncol(W)<3) && (PCA))
stop("Error: You choose PCs for reconstruction by pool of potential predictors N<3)")

# Time plot of predictand 
tit1 <- paste('Winter Precipitation')
plot(yrv,v,type='b',xlab='Year',ylab='Precipitation (mm)',main=tit1,col='blue')
abline(h=mean(v),col='gray')

# Listing of variable codes in lagged chronology matrix
# sprintf('%s','Variables (cols) in the lagged chronology matrix W, before any correlation screening')
# sprintf('%s',colnames(W))
rm(tit1)
```

################################################################ 

## PART 2. Correlation screening of lagged chronology matrix

################################################################# 

```{r}

#--- Cull common period predictand and all cols of lagged chron matrix
B<-cbind(yrv,v)
A<-cbind(yrW,W)
D<-PeriodCommon(A,B) # D$Y, D$X: predictand predictor matrices, with year in col 1
mv1<-D$tsp-D$tgo+1 # number of observations in common period
v1<-D$Y[,2] # common-period predictand
yrv1<-D$Y[,1] # common-period time vector
ntemp<-ncol(D$X) # col-size of lagged predictor matrix, with year col
W1<-D$X[,2:ntemp] # common period predictor matrix
W1<-as.matrix(W1) # in case 1-col W1
yrW1<-yrv1 # common period time vector, duped
colnames(W1)<-colnames(W) # column names remain the same as before culling 

#--- If you specified yrgoc or yrspc, truncate W1, yrW1, v1, yrv1 accordingly
# If you specified a yrgoc or yrspc outside the range of yrv1, give error message
Lfront=FALSE
Lback=FALSE

# front
Ltemp = is.na(yrgoc)
if (Ltemp){
  yrgoc <- yrv1[1]
} else {
  if (yrgoc<yrv1[1]){
    strtemp<-paste('Re-specify yrgoc.\n','You set yrgoc = ',sprintf('%5.0f',yrgoc),'.',
                   '\nThe data are not consistent with yrgoc<',sprintf('%5.0f',yrv1[1]))
    stop(strtemp)
  } else {
  }
  if (yrgoc == yrv1[1]) {
  } else {
    Lfront=TRUE
  }
}

# back
Lstopc <- FALSE # initial flag for locking in stop year of calibration to what user specified, as yrspc
Ltemp <- is.na(yrspc)
if (Ltemp){
  Lstopc <- TRUE
  yrspc <- yrv1[length(yrv1)]
} else {
  if (yrspc>yrv1[length(yrv1)]){
    strtemp<-paste('Re-specify yrspc.\n','You set yrspc = ',sprintf('%5.0f',yrspc),'.',
                   '\nThe data are not consistent with yrsp > ',sprintf('%5.0f',yrv1[length(yrv1)]))
    stop(strtemp)
  } else {
  }
  if (yrspc == yrv1[length(yrv1)]) {
  } else {
    Lback=TRUE
  }
}

# If you have specified a calibration period shorter than the common period of lagged chronologies and predictand, 
# truncate accordingly because want the correlation screening as well as model calibration on the shortened interval
Ltemp = Lfront || Lback
if (Ltemp){
  L <- (yrv1 >= yrgoc) & (yrv1<=yrspc)
  v1 <- v1[L]; yrv1 <- yrv1[L]
  W1 <- W1[L,]; yrW1 <- yrW1[L]
  mv1 = length(v1)
} else {
}
rm(ntemp,A,B,D,Lfront,Lback,Ltemp,L)


# Check specified years for "independent verification" (IV)." If both yrgoi and yrspi are NA, no
# IV is attempted. Settings yrgoi and yrspi must BOTH be years or BOTH be NA. If IV is desired, 
# no overlap is allowed of period yrgoi-yrspi with yrgoc-yrspc, and the IV period must be at least 5 years long. 
LIV <- !(is.na(yrgoi) && is.na(yrspi)) # if true, you want IV 
if (sum(is.na(c(yrgoi,yrspi)))==1){
  stop('Not allowed to set just one of yrgo1 and yrspi to NA')
}
if (LIV){
  # IV period must be at least 5 years
  nIV <- yrspi-yrgoi+1
  if (nIV < 5){
    stop('Specified period for independent verification shorter than 5 years')
  } else {
    yrIV <- yrgoi:yrspi # vector of IV years
  }
  # IV period cannot overlap calibration/validation period
  if (any(yrIV %in% yrv1)){
    stop('Specified period for indpendent verification overlaps calibration period')
  }
}


# Status: yrv1,v1, yrW1, W1 are calibration-period predictand (v1) and
# predictor. Names of columns of W1, colnames(W1), are coded to indicate lag

#--- Correlate v1 with cols of W1 and identify threshold correlation,'
# rc, marking the floor(f*mv1) highest correlation, where f is a decimal fraction and
# mv1 is the number of observations in the calibraiton period. 

#--- Critical threshold absolute correlation
nkeep<-floor(f*mv1) # want no more than this number of potential predictors in pool

# Cannot keep more cols of W1 than W1 has
nkeep <-min(nkeep,ncol(W1))

r<-cor(W1,v1) # Pearson correlation
rs<-sort(abs(r),decreasing=TRUE,index.return=TRUE) # sorted r (list object)
rc = rs$x[nkeep] # threshold absolute correlation for screening

# Pull subset of predictors with absolute correlation >rc
j<-rs$ix;
nW2 <- nkeep # number of screened potential predictors
ix1<-j[1:nkeep] # column indices of screened ...
W2<-W1[,ix1] # correlation-screened set of predictors; note these are ordered
# left to right in W2 in order of decreasing absolute correlation with v1
W2<-as.matrix(W2)
yrW2<-yrW1
# STATUS: W2, yrW2 are the screened potential predictors; ix1 is
# pointer to columns of the screened potential predictors in the
# full matrix of lagged potential predictors

# Barplot of calib-period correlation of predictand with lagged chronologies
barplot(t(r))
tit1 <- paste('r of predictand with lagged chronologies ',
              as.integer(yrgoc),'-',as.integer(yrspc),'\n',as.integer(nkeep),' passed screening',
              '(critical absolute r=',sprintf('%5.2f',rc),')')
title(tit1)

# List tree-ring variables passing screening
sprintf('%s','Variables passing screening (these in W2 in order L to R by ranked absolute r)')
sprintf('%s',colnames(W2))




```

################################################################ 

## PART 3. Optionally extend calibration period

################################################################# 

```{r}
Wtemp<-W[,ix1] # full lagged matrix, screened cols only
Wtemp<-as.matrix(Wtemp)
L <- yrW>=yrv1[1] & yrW<=yrv1[mv1] # logical to years of predictand
Wtemp<-Wtemp[L,]
yrWtemp<-yrW[L]
Wtemp<-as.matrix(Wtemp)

#--- Revise screened predictor pool calibration years if possible and your input setting of yrspc was NA
if ((nrow(Wtemp) != nrow(W2)) && Lstopc ) {
  W2<-Wtemp
  yrW2<-yrWtemp
  # Check yrv against yrW2 and set calibration period as years with complete
  # cases of v and W2
  B<-cbind(yrv,v)
  A<-cbind(yrW2,W2)
  D<-PeriodCommon(A,B) # D$Y, D$X: predictand predictor matrices, with year in col 1
  mv1<-D$tsp-D$tgo+1 # number of observations in common period
  v1<-D$Y[,2] # common period predictand
  yrv1<-D$Y[,1] # common period time vector
  ntemp<-ncol(D$X) # col-size of lagged predictand matrix, with year col
  W2<-D$X[,2:ntemp] # common period predictand matrix
  yrW2<-yrv1 # common period time vector, duped
  colnames(W2)<-colnames(W)[ix1] # column names remain the same as before culling 
  yrspc=yrv1[length(yrv1)] # revised end year for calibration
  rm(ntemp,A,B,D)
}
```

################################################################################ 

## PART 4. PCA transformation of chronologies

############################################################################### 

```{r}
#--- Get matrix of screened lagged chronologies for full common period of those
# chronologies. This will be used for the PCA, and will also be needed later for
# reconstruction by the non-PCA option
Wf<-W[,ix1] # full matrix of screened lagged chronologies
yrWf <-yrW
L<-complete.cases(Wf) # logical to years with no missing data
W3<-Wf[L,] # common-period screened lagged Chronologies
yrW3<-yrWf[L]

# PCA: option activated only if 3 or more screened lagged chronologies and more than one chronology in input tree-ring matrix
if (dim(W3)[2]>2){
# if (dim(W3)[2]>2 && dim(U)[2]>1){ # earlier version also had condition on number of original chronologies (before any lagging) for PCA to be attempted
  resPCA<-princomp(x=W3,cor=TRUE,fix_sign=TRUE) # PCA, on correlation matrix
  summary(resPCA) # summary of PCA
  screeplot(resPCA) # scree plot of variance accounted for by PCa
  titTemp <- paste('Scree plot of principal components of screened lagged chronologies',
                   '\nFor details of the PCA, type summary(resPCA) at R console')
  title(titTemp)
  
  A<-resPCA$scores  # the new times series; PC "scores"
  yrA<-yrW3
  plot(yrA,A[,1],type="l",xlab="Year",ylab="Score") # time plot of PC1
  title('Scores of PC1 of screened lagged chronologies')
  
  
  # Pull rows of A for regression modeling; 
  yrA2<-intersect(yrA,yrv1)
  L<- (yrA>=yrA2[1]) & (yrA<=yrA2[length(yrA2)])
  A2<-A[L,]
  
  # STATUS. resPCA is a "princomp" object. A summary of the PCA is
  # printed by >summary(resPCA). Loading, scores and other data
  # can be obtained by such commands as pcW3$loadings. For 
  # regression and reconstruction, the scores are stored
  # in matrix A2 and A, with years in yrA2 and yrA
} else {
  resPCA <- ('Too few tree-ring variables to justify PCA.')
}
```

################################################################################ 

## PART 5. Regression modeling

################################################################################ 

```{r}
# Review of needed variables, which are ready:
#   v1, yrv1:  [numeric] time series of predictand and its years
#   W2, yrW2: [matrix, vector]: time series matrix of screened predictor pool
#   A2, yrA2 is the PC equivalent of W2, yrW2
#   colnames(W2) [character] ids of screened lagged predictors, coded by column in
#     original loaded tree-ring matrix and by lag
#   ix1[integer]: index of lagged screened predictors indicating the columns 
#     those in the matrix of the lagged tree-ring matrix before screening. 
#     ix1 will be needed later when we want to pull tree-ring data to generate
#     the long-term reconstruction
#
#   For that and for reconstruction we have the full lagged tree-ring matrix
#     W, yrW [matrix, numeric], and we know that W1 is the sub-matrix of
#     columns ix1 of W


if (!PCA | (dim(W3)[2]<2)){
  # Regression on chronologies themselves (possibly with lags)
  namesW2<-colnames(W2) # want these names as a separate input arg to ForwStep1
  F<-ForwStep1(W2,namesW2, v1,kStopWhen,nNeg,nPos) # forward stepwise 
  #   regression of v1 on W2
}else{
  namesA2<-colnames(A2) # names of PCs: Comp.1, Comp.2, etc
  F <- ForwStep1(A2,namesA2,v1,kStopWhen,nNeg,nPos) # forward stepwise 
  #   regression of v1 on A2
} 


#--- Get the model object see a summary 
M<-F$Model
summary(M)

# Get fitted values, bind into a matrix with the observed, and plot
# time series
yh <- M$fitted.values
yrgo<-yrv1[1]
Z<-cbind(v1,yh)
Z.ts <-ts(Z,start=yrv1[1])
ts.plot(Z.ts,type='b',col=c('blue','red'),xlim=c(yrgo,yrv1[length(v1)]), xlab="Year", ylab="Precipitation (mm)")
abline(h=0)
#--- Add a title and line at the mean
title('Observed (blue) and Reconstructed (red)')
abline(h=mean(v1),col='grey')

# Scatterplot of predicted on observed, with least-squares
# fit straight lin as well as loess fit of locattion an spread
scatterplot(v1,yh, xlab='Observed',ylab='Predicted')
title(sprintf('Adjusted R-squared = %5.2f',F$RsquaredAdj))
```

############################################################################### 

## PART 6. Reconstruction

################################################################################ 

```{r}
#X<-as.data.frame(W3[,F$ColsInModel]) # predictor matrix f
if (!PCA){
  X<-W3[,F$ColsInModel] # predictor matrix for recon
  yrX<-yrW3
}else{
  X<-A[,F$ColsInModel] # predictor matrix for recon
  yrX<-yrA
}
X <- as.matrix(X)
mX = nrow(X) # number of rows (years) in predictor matrix
Xones=as.matrix(rep(1,mX)) 
X<-cbind(Xones,X) # add column of 1's to predictor matrix
b<-F$Coefficients # regression coeffs
b<-as.matrix(b) #... to row vector
v1hat<-X %*% b   # reconstructions

#-- renaming for simplicity; know have v1,yrv as calib-period predictand
y <- v    # observed predictand, all available years
yry <- yrv
yc <- v1; # calibration period predictand
yryc <-yrv1
yh <- v1hat # reconstructed ("hat") predictand
yryh <-yrX 

#---- Quality control: are predictions for calib period the same as
# fitted values from the lm function?
yf<-fitted(M) # fitted valued from model object
yryf <-yrv1 # these are for the calib years
yr1 <- yryf[1] # start year of calib
yr2 <-yryf[length(yryf)] # end year of calibration

L <- yryh>=yr1 & yryh<=yr2
ycheck<-yh[L]

# Check that specified period yrgo-yrspi for indpendent verification consistent with
# overlap of reconstructed and observed predictand 
# Know that if LIV =TRUE, you want independent verification, and if FALSE, will skip it
if (LIV){
  yrgoO <- max(c(yrv[1],yryh[1])) # first year of overlap of recon and observed predictand
  yrspO <- min(c(yrv[length(yrv)]),yryh[length(yryh)]) # last year of ...
  yrO <- yrgoO:yrspO
  Lkill <- !all(yrIV %in% yrO)
  if (Lkill){
    str_kill <- paste('You specified independent verification (IV) period ',as.integer(yrgoi),'-',
                      as.integer(yrspi),' but overlap of observed and reonstructed predictand covers only ',
                      as.integer(yrgoO),'-',as.integer(yrspO),
                      '. And keep in mind that IV cannot include any years in the ',as.integer(yrgoc),
                      '-',as.integer(yrspc),' calibration period.')
    stop(str_kill)
  }
  rm(Lkill)
}


# Quality control (QC1)
#--- Print 2-col matrix and check that the column of lm-fitted values is 
# identical to the column of reconstructed values generated by applying the
# regression equation to the matrix of screened lagged chronologies for
# the calibration period
QC1<-cbind(as.matrix(yf),as.matrix(ycheck))
print(QC1)
```

################################################################################ 

## PART 7. Independent verification

```{r}

LIV  <- is.na(yrgoi) | is.na(yrspi)
if (!LIV){
# Recall have reconstruction in yh, yryh; observed predictand in y,yry; calibration predictand in yc
# Will call function NashSutt to do the independent verification

kplotIV <- 1 # plot option calling for plot in next device
outputDir <- NA   # ignored by NashSutt for kplotIV=1
gFileType <- 'png' # file-type of plot (ignored for kplotIV=1)

# Cull time series of observed and reconstructed predictand for IV period
L <- (yry >= yrgoi) & (yry <= yrspi)
yIV  <- y[L]; yryIV <- yrIV # observed
L <- (yryh >= yrgoi) & (yryh <= yrspi)
yhIV  <- yh[L]; yryhIV <- yryh[L] # recon
ycMean  <- mean(yc)

ResIV <- NashSutt(yIV, yhIV, yrIV, ycMean, kplotIV, outputDir,gFileType) 
} else {
  ResIV <- 'You elected to not do independent verification'
  
}




```

################################################################################ 

## PART 8. Analysis of residuals

################################################################################ 

```{r}

# A few tests on residuals
F$DurbinWatson <- durbinWatsonTest(M) #--- Durbin-Watson test for lag-1 autcorrelation 
hLillie <- lillie.test(M$residuals); # Lilliefors test for normality


#--- Time plot with Mann-Kendall test for trend
  # Prepare input for mannken1
  X <-cbind(yryc,M$residuals) # matrix with year and regression residuals

  kopt<- c(2,1) # want plot; want adjustment of significance of Mann-Kendall statistic
  # for autocorrelation if warranted
  kplot <-1 # non-TRISH version, simple plot with trend line
  ylabTemp1 <-'Residual'
  ylabTemp2 <- 'Detrended Residual'
  textPlot <- c('Regression Residuals with Nonparametric-Fit Trend Line,','Year',ylabTemp1,
                ylabTemp2)
  FigNumber=0 # meaningless inpiut to mannken1 outside of TRISH
  Din <- list(X=X,kopt=kopt,kplot=kplot,NextFigNumber=FigNumber,textPlot=textPlot,outputDir='Null')
  
  # mannken1 to get statistics and plot
  ResMK <- mannken1(Din)
  rm(Din)
  
  #---Histogram, with annotated Lilliefors test  p-value
  Tit1 <- paste('Residuals, E',
                '\n  (p=',sprintf('%.2g',hLillie$p.value),'for H0 that E normal, from Lilliefors test)')
  hist(M$residuals,xlab='E (predictand data units)',ylab='Frequency',main=Tit1)
  
  # acf and its 95% CI
  len1 <- floor(length(yryc)/4)
  lagsPlot <- min(len1,20)
  acfMy <- acf(M$residuals, lag.max=lagsPlot, type = "correlation",
               plot = TRUE,xlab='Lag (yr)', ylab = 'ACF')
  r1 <- acfMy$acf[2] # lag-1 autocorrelation

      #               '\n  (p=',sprintf('%.2g',BP$p),' for H0 that E homoscedastic)',
    #               '\n(from Breusch-Pagan Test)')
  r1Str =paste('r(1)=',sprintf('%5.2f',r1),'\nDW  statistic=',sprintf('%6.1f',F$DurbinWatson$dw),
               '\np=',sprintf('%6.3f',F$DurbinWatson$p))
  text(lagsPlot,0.9,r1Str,adj=c(1,1))
  
  
  # Scatter of residuals against fitted values, with annotated p=value for Breusch-Pagan test. 
  BP <- ncvTest(M)
  if (BP$p>0.05){
    strBP <- ' (cannot reject at 0.05)'
  } else {
    strBP <-' (reject at 0.05)'
  }
  Tit2 <- paste('Breusch-Pagan test; H0: errors homoskedastic',
                '\np=',sprintf('%G',BP$p),strBP)

png("figs/Breusch-Pagan.png", width = 750 , height = 420)
  plot(M$fitted.values,M$residuals,xlab='Fitted Values',
       ylab='Residuals',
       main=Tit2)
  abline(h=0,lty=2,col='#808080') # dash gray
dev.off()

```

################################################################################ 

## PART 9. Split-sample analysis to check stability of reconstruction model

################################################################################ 

```{r}
# Review
# M$Model is data frame with calibration-period  predictand in col 1 and predictors in remaining columns 
# yrv1 is vector of predictand years for calibration period, with length mv1
# Need to regress col 1 of M#Model on the remaining columns fo#
# Will call Meko's function ssValid to do the split-sample calibration-validation
#   First make pointer to the rows of M$Model corresponding to A (early) and B  (late)
#   Call ssValid twice, switching the row pointer for calibration and valdation
#   Store results for both halves.
# Plot results in 1x1 figure with time plots of observed and validation-predicted predicand, with vertical line splitting halves

# Make pointer to halves
nE  <- ceil(mv1/2) # sample size, early half
nL = mv1-nE  # sample size late
i1 <- 1:nE; i2 <- (nE+1):mv1   # pointer to rows

y1 <- as.matrix(M$model[,1])
X1 <- as.matrix(M$model[,-1])
nX1 <- dim(X1)[2]
iuse <- 1:nX1

resSScalE <- ssValid(y1,X1,i1,i2,iuse) # cal on first, val on 2nd
resSScalL <- ssValid(y1,X1,i2,i1,iuse) # cal on first, val on 2nd

atemp <-bquote (.R^2)
strE <- paste('RE=',sprintf('%5.2f',resSScalL$RE))
strL <- paste('RE=',sprintf('%5.2f',resSScalE$RE))

# Get SS predctions
yra1 <-yrv1[i1]; yra2 <-yrv1[i2]
yrmid <- 0.5*(yra1[length(yra1)] + yra2[1])
a1 <-resSScalL$ssPreds
a2 <-resSScalE$ssPreds

# x and y for null prediction horiz lines
mn1 <- resSScalL$MeanObsCalibPd # Late period obs calib mean
mn2 <- resSScalE$MeanObsCalibPd # Early period obs calib mean
xmn1 = c(yra1[1],yra1[length(yra1)])
xmn2 = c(yra2[1],yra2[length(yra2)])
ymn1 =c(mn1,mn1); ymn2 =c(mn2,mn2)

# build ylims
fscale1 <- 0.01; fscale2 <- 0.6
yhi <- max(y1,a1,a2); ylo <- min(y1,a1,a2)
deltay <- (yhi-ylo)
yhi <- yhi+fscale2*deltay
ylo <- ylo - fscale1*deltay
ylims <-c(ylo,yhi)

xlims <- c(yrgoc-1, yrspc+1)
plot(yrv1,v1,type="o", col="black", pch="o", 
     xlab='Year',ylab="Predictand", lty=1,xlim=xlims,ylim=ylims)
lines(yra1, a1, col="red",lty=2, pch="*",type="b")
lines(xmn1,ymn1,col="red")
lines(yra2, a2, col="blue",lty=2,pch="*",type="b")
lines(xmn2,ymn2,col="blue")

abline(v=yrmid,col="gray")
text(yra1[1],ylims[2],strE,adj=c(0,1))
text(yra2[1],ylims[2],strL,adj=c(0,1))
title('Split-sample prediction skill')

legend("topright", legend=c("Observed","Recon-early","Null-early","Recon-Late","Null-late"),
       col=c("black","red", "red","blue","blue"),
       lty=c(1,2,1,2,1), cex=0.8)

whatsUp <- 'Results for split-sample (SS) analysis, calibrated on early (...calE) and late (...calL)' 
EarlyHalf <- paste('Early = ',as.integer(yra1[1]),'-',as.integer(yra1[length(yra1)]))
LateHalf <- paste('Late = ',as.integer(yra2[1]),'-',as.integer(yra2[length(yra2)]))
resSS <- list(what=whatsUp,EarlyHalf=EarlyHalf,LateHalf=LateHalf,
               resSScalE=resSScalE,resSScalL=resSScalL)
rm(whatsUp)

```

################################################################################ 

## PART 10. Confidence interval for reconstruction

############################################################################### 

```{r}
#--- Compute calibration root-mean-square error (RMSEc), just to be able to compare 
# that to the cross-validation RMSE.
# reconstructed value, lower 50% CI and upper 50% CI; then plot the the series
RMSEc <-  sqrt(sum(M$residuals^2)/(M$df.residual)) # root-mean-square error of calib
RMSE <- F$RMSEcv # but use the cross-validation RMSE for the confidence interval on reconstruction
yhlo<-yh-0.6475*RMSE # upper 50% 
yhhi<-yh+0.6475* RMSE # lower 50%
YhCI<-cbind(yh,yhlo,yhhi)
namestemp<-c('Recon','Lower 50%','Upper 50%')
YhCI.ts<- ts(YhCI,start=yryh[1],frequency=1,names=namestemp)

# time series for plot and CI
w <- yh; yrw <- yryh #  reconstruction 

# Compute shaded polygon x and y
Xtemp <- cbind(yryh,yhlo,yhhi) # matrix with year as col 1, lower CI as col 2, upper CI as col 3
ResTemp <- xyCI(Xtemp)
xP <- ResTemp$x;   yP <- ResTemp$y

# Limits for plot
yLo <- min(Xtemp[,2:3])
yHi <- max(Xtemp[,2:3])
ynudge <- 0.02 * (yHi-yLo)
ylims = c(yLo-ynudge, yHi+ynudge)
xlims = c(yryh[1]-1,(max(yryh)+1))

# Strings for plot
strRecYrs <- paste(sprintf('%d',yryh[1]),'-',sprintf('%d',yryh[length(yryh)]),sep='')
Tit1 <- paste('Reconstruction',', ',strRecYrs,
              '\n(50% CI shaded; dashed line at reconstructed mean)',sep='')
ylab1 <- 'Predictand'

# Plot
plot(yryh,yh,type="l",col="blue",xlim=xlims,ylim=ylims,
     ylab=ylab1,xlab='Year',main=Tit1)
abline(h=mean(yh),lty=2,col='#808080') # dash gray
polygon(xP,yP, col=rgb(1.00,0,0,0.1),border=NA) # mustard

# # OLD CODE, COMMENTED OUT
# YhCI<-cbind(yh,yhlo,yhhi)
# namestemp<-c('Recon','Lower 50%','Upper 50%')
# YhCI.ts<- ts(YhCI,start=yryh[1],frequency=1,names=namestemp)
# ts.plot(YhCI.ts,col=c('black','red','red'),xlim=c(yryh[1],yryh[length(yh)]))
#  
# #% add line at mean, and title
# abline(h=mean(yh),col="blue",lty=3)
# title('Reconstruction with 50% CI')


#--- STATUS
# yh [matrix] is reconstruction for years yryh [numeric]
# M [lm object] returned model from lm function
# F$Rsquared, F$RsquaredAdj, etc, model statistics stored in list F. The list
#   items actually generated from lm can also be extracted from the model 
#   object M. Other data and statistics can also be pulled from that model
#   object. For example summary(M), fitted(M), ... 
#   the lis
# colnames(W2)[F$ColsInModel] ... this returns "ids" of variables in the model.
#   "Xn" refers to column n of chronology in input data frame; Ni or Pi 
#   refer to lag i, negative (N) or postive (P) relative to year of predictand
```


############################################################################### 

## PART 11. New plots

############################################################################### 

```{r}
# Colors
KGU.std <- "#2E5F7F"
KGU.res <- "#DA7842"
NB.std <- "#35291D"
NB.res <- "#56402D"
precip <- "#5A84D4"
prdctnd <- "#CB4E47"
split.early <- "#EF8636"
split.late <- "#8D69B8"
```
```{r}
# Chronologies
alph <- 0.7
chrons <- read.csv(here("chrons.csv"))

chron.std <- ggplot(data = chrons) +
  geom_path(aes(x = Year, y = NB.std), color = NB.std) +
  geom_path(aes(x = Year, y = KGU.std), color = KGU.std) +
  xlab(NULL) +
  ylab("RWI") +
  theme_par() +
  theme(plot.margin = margin(0,0,0,0))
chron.res <- ggplot(data = chrons) +
  geom_path(aes(x = Year, y = NB.res), color = NB.res) +
  geom_path(aes(x = Year, y = KGU.res), color = KGU.res) +
  ylab("RWI") +
  theme_par() +
  theme(plot.margin = margin(0,0,0,0))
chron.stack <- chron.std / chron.res

png("figs/stack.png", width = 1200 , height = 500)
plot(chron.stack)
dev.off()


# chron.stack <- chrons |> ggstackplot(
#   x = Year, y = c(NB.std, KGU.std, NB.res, KGU.res),
#   overlap = (c(1, 0, 1)),
#     template = ggplot() +
#     theme_stackplot() +
#     labs(x = "Year", y = "RWI"),
#   simplify_shared_axis = TRUE,
#   add = list(
#     geom_path(color = NB.std, alpha = alph) +
#       scale_y_continuous(guide = "none"),
#     geom_path(color = KGU.std, alpha = alph) + 
#       labs(title = "Standard"),
#     geom_path(color = NB.res, alpha = alph)+
#       scale_y_continuous(guide = "none"),
#     geom_path(color = KGU.res, alpha = alph) + 
#       labs(title = "Residuals")
#   )
# )
```
```{r}
# Winter precip
png("figs/winter-precip.png", width = 750 , height = 420)
tit1 <- paste('Winter Precipitation (Previous October—April)')
plot(yrv,v,type='b',xlab='Year',ylab='Precipitation (mm)',main=tit1,col=precip)
abline(h=mean(v),col='gray')
dev.off()
```


```{r}
# Observed and reconstructed
png("figs/obsv-recon.png", width = 750 , height = 420)
ts.plot(Z.ts,type='b',col=c(precip,prdctnd),xlim=c(yrgo,yrv1[length(v1)]), xlab="Year", ylab="Precipitation (mm)")
abline(h=0)
title('Observed (blue) and Reconstructed (red)')
abline(h=mean(v1),col='grey')
dev.off()
```

```{r}
# Split-sample validation
xlims <- c(yrgoc-1, yrspc+1)
fscale1 <- 0.01; fscale2 <- 0.6
yhi <- max(y1,a1,a2); ylo <- min(y1,a1,a2)
deltay <- (yhi-ylo)
yhi <- yhi+fscale2*deltay
ylo <- ylo - fscale1*deltay
ylims <-c(ylo,yhi)

png("figs/split-sample.png", width = 750 , height = 420)
plot(yrv1,v1,type="o", col="black", pch="o", 
     xlab='Year',ylab="Precipitation (mm)", lty=1,xlim=xlims,ylim=ylims)
lines(yra1, a1, col=split.early,lty=2, pch="*",type="b")
lines(xmn1,ymn1,col=split.early)
lines(yra2, a2, col=split.late,lty=2,pch="*",type="b")
lines(xmn2,ymn2,col=split.late)
abline(v=yrmid,col="gray")
text(yra1[1],ylims[2],strE,adj=c(0,1))
text(yra2[1],ylims[2],strL,adj=c(0,1))
title('Split-sample prediction skill')
legend("topright", legend=c("Observed","Recon-early","Null-early","Recon-Late","Null-late"),
       col=c("black",split.early,split.early,split.late,split.late),
       lty=c(1,2,1,2,1), cex=0.8)
whatsUp <- 'Results for split-sample (SS) analysis, calibrated on early (...calE) and late (...calL)' 
EarlyHalf <- paste('Early = ',as.integer(yra1[1]),'-',as.integer(yra1[length(yra1)]))
LateHalf <- paste('Late = ',as.integer(yra2[1]),'-',as.integer(yra2[length(yra2)]))
resSS <- list(what=whatsUp,EarlyHalf=EarlyHalf,LateHalf=LateHalf,
               resSScalE=resSScalE,resSScalL=resSScalL)
dev.off()
```

```{r}
## Reconstruction and PRISM

# plot(yryh,yh,type="l",col="blue",xlim=xlims,ylim=ylims,
#      ylab=ylab1,xlab='Year',main=Tit1)
# abline(h=mean(yh),lty=2,col='#808080') # dash gray
# polygon(xP,yP, col=rgb(1.00,0,0,0.1),border=NA) # mustard

reconst.precip <- as.data.frame(cbind(yryh, yh)) |>
  mutate(Year = as.numeric(yryh), 
    Precip = as.numeric(yh)) |>
  select(c(Year, Precip))
reconst <- ggplot(data = NULL,) + 
  geom_rect(aes(xmin = 1480, xmax = 1594, ymin = -Inf, ymax = Inf), col = "#E8E0D9", alpha = 0.2) +
  geom_hline(aes(yintercept = mean(yh)), col = "#808080") +
  geom_ribbon(aes(x = yryh, ymin = yhlo, ymax = yhhi), col = prdctnd, alpha = 0.3) +
  geom_path(aes(x = yryh, y = yh), col = prdctnd, alpha = alph, linewidth = 0.5) +
  geom_path(aes(x = yrv, y = v), col = precip, alpha = alph, linewidth = 0.5) + 
  labs(x = "Year", y = "Precipitation (mm)", title = "Reconstruction, 1480—2020") +
  theme_par()
png("figs/reconst.png", width = 750 , height = 500)
plot(reconst)
dev.off()
```

############################################################################### 

#### PART 12. Save reconstruction and statistics

############################################################################### 

```{r}
# Edit following line so that path matches where you want the .RData output file
save(yrgoc,yrspc,YhCI.ts, Z.ts,F,resSS,resPCA,file=paste(here(), 'RecOutput.RData', sep = ""))
```


